Limited physiological plasticity and reduced relative variance in physiological rates of ectotherm populations under climate change

### 2024-07-04 ms.qmd
Removing the ecological consequences section from discussion and moving a few sentences to the intro. I think some of these are better placed there anyway.

@@

We show that acclimation of physiological rates and changes in relative variance differ across habitats. Our findings confirm previous results that quantify the different capacity of terrestrial, marine and freshwater ectotherms to acclimate [@Seebacher2015-wi; @Gunderson2015-hz; @morley2019physiological]. Our analysis generally supports that of @Seebacher2015-wi, @Gunderson2015-hz and @morley2019physiological that all show a general inability of terrestrial ectotherms to physiologically acclimate. The change in acclimation $Q_{10}$ we found was similar to @Seebacher2015-wi for freshwater organisms (~17%), but higher in marine ectotherms (decrease of 16% versus ~10% in @Seebacher2015-wi), and opposite in terrestrial ectotherms (increase of ~6% compared to an ~8% decrease in @Seebacher2015-wi), possibly due to our use of new $Q_{10}$ based effect sizes that allow us to control for sampling variance. Importantly, while marine and freshwater ectotherms were capable of partial acclimation, the effect size was small (amounting to $Q_{10}$ dropping from 1.8 to 1.6), suggesting acclimation will, on average, provide limited capacity to maintain physiological rates in the face of climate change.

Changes in relative variance in physiological rates across habitats was nearly equivalent (marine) or larger (terrestrial) compared to their capacity to acclimate. Only in freshwater ectotherms did acclimation in physiological rates coincide with no change in relative variation. Mechanistically, it is unclear what is contributing to reduction in relative variation in physiological rates at higher temperatures. One possible explanation may be that biochemical reactions are operating more efficiently or are relying on more conformationally stable enzymes at higher temperatures [@tattersall2012coping; @angilletta2009thermal; @somero1995proteins]. Such a mechanism may also help explain the differences in relative variance across habitat types. Terrestrial ectotherms may be relying on more specialised enzymatic variants that operate over a more limited temperature range individually but together allow for organisms to function over a wider thermal range. Needless to say, our results further highlight the potential vulnerability of terrestrial ectotherms to climate change, while calling into question if acclimation in marine systems will be able to counteract the potentially negative effects of reductions in relative variance. Assuming changes in relative variation in physiological rates are underpinned by genetic variation, and that there is a genetic correlation with fitness, such reductions in physiological variance could limit a populations ability to adapt to future climates [@urban2023and].

Importantly, while we seen a reduction in relative variation, absolute variation in physiological rates did increase with temperature (@fig-s1) -- albeit to a much lesser extent in terrestrial and marine habitats compared to freshwater -- supporting the general notion that capacity for adaptation may still be limited. As of yet, it is unclear whether a focus on relative or absolute variance should be prioritized or how behavioural strategies of ectotherms might be able to compensate for the general weak acclimation abilities of many ectotherms. For example, terrestrial species have greater opportunity for behavioural microhabitat selection which is expected to decrease the impacts of climate change in thermally heterogeneous environments [@huey2012predicting]. Future theoretical models, that vary both mean and variance, are needed to better understand the broader evolutionary and ecological implications of our findings.

@@

### ***Ecological consequences of reduced relative variation in physiological rates***
Relative variance reductions in metabolic rates and performance (e.g., sprint speed) were most strongly impacted by increased temperature. Changes in variability of traits governing energy demand could have important consequences on the flow of energy within and between populations, communities, and ecosystems [@hendry2016key; @seebacher2023can; @barneche2021warming; @Sanderson2023]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key; @hart2016variation]. Maintaining intrapopulation variability in physiological rates in a warmer world may therefore be important for population resilience to climate change. For example, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population fluctuations compared to scenarios with no variation [@kooijman1989population]. Our results suggest that terrestrial ectotherms will likely face greater challenges to adapting to climate change [@hoffmann2013upper; @Gunderson2015-hz; @Pottier2022], leading to greater rates of extinction and reduced productivity in terrestrial habitats. However, data testing the impacts of changes in variance on populations are limited to very few taxa. Future experimentation across a broader range of taxa should manipulate levels of variance to understand its ecological outcomes.

>>>>

Changes in variability of traits governing energy demand could have important consequences on the flow of energy within and between populations, communities, and ecosystems [@hendry2016key; @seebacher2023can; @barneche2021warming; @Sanderson2023]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key; @hart2016variation]. Maintaining intrapopulation variability in physiological rates in a warmer world may therefore be important for population resilience to climate change.

@@

We show that acclimation of physiological rates and changes in relative variance differ across habitats. Our findings confirm previous results that quantify the different capacity of terrestrial, marine and freshwater ectotherms to acclimate [@Seebacher2015-wi; @Gunderson2015-hz; @morley2019physiological]. Our analysis generally supports that of @Seebacher2015-wi, @Gunderson2015-hz and @morley2019physiological that all show a general inability of terrestrial ectotherms to physiologically acclimate. The change in acclimation $Q_{10}$ we found was similar to @Seebacher2015-wi for freshwater organisms (~17%), but higher in marine ectotherms (decrease of 16% versus ~10% in @Seebacher2015-wi), and opposite in terrestrial ectotherms (increase of ~6% compared to an ~8% decrease in @Seebacher2015-wi), possibly due to the use of new $Q_{10}$ based effect sizes that allow us to control for sampling variance. Importantly, while marine and freshwater ectotherms were capable of partial acclimation, the effect size was small (amounting to $Q_{10}$ dropping from ~1.8 to 1.6).

Changes in relative variance in physiological rates across habitats was nearly equivalent (marine) or larger (terrestrial) compared to their capacity to acclimate. Only in freshwater ectotherms did acclimation in physiological rates coincide with no change in relative variation. Mechanistically, it is unclear what is contributing to the reduction in relative variation in physiological rates at higher temperatures. One possible explanation may be that biochemical reactions are operating more efficiently or are relying on more conformationally stable enzymes at higher temperatures impacting between individual variation [@tattersall2012coping; @angilletta2009thermal; @somero1995proteins]. Such a mechanism may also help explain the differences in relative variance across habitat types. For example, terrestrial ectotherms may rely on more specialised enzymatic variants that operate over a more limited temperature range individually but together allow for organisms to function over a wider thermal range necessary to operate in more variable thermal environments [@angilletta2009thermal]. Acclimation responses in marine and freshwater organisms would also result in the maintenance, or even increase, in variance at higher temperatures due to the expression of alternative enzymes that function more effectively at higher temperatures. Given that biochemical traits in marine and freshwater organisms acclimated, and their relative variance increased in freshwater after acclimation, this seems plausible.

Our results further highlight the potential vulnerability of terrestrial ectotherms to climate change, while calling into question if acclimation in marine systems will be able to counteract the potentially negative effects of reductions in relative variance. Assuming changes in relative variation in physiological rates are underpinned by genetic variation, and there is a genetic correlation with fitness, such reductions in physiological variance could limit a populations ability to adapt to future climates [@urban2023and]. Of course, considering how behavioural modifications buffer organisms from high temperatures is also needed. For example, terrestrial species have greater opportunity for behavioural microhabitat selection which is expected to decrease the impacts of climate change in thermally heterogeneous environments [@huey2012predicting]. Importantly, while we seen a reduction in relative variation, absolute variation in physiological rates did increase with temperature (@fig-s1) -- albeit to a much lesser extent in terrestrial and marine habitats compared to freshwater -- supporting the general notion that capacity for adaptation may still be limited in terrestrial and marine habitats. As of yet, it is unclear whether a focus on relative or absolute variance should be prioritized. Future theoretical models, that vary both mean and variance, are needed to characterise the interplay between changes in mean and variance.

@@
